---
title: "2026 Ecosystem Summary Report pilot for Region 4X"
author: "Maritimes EBFM WG"
date: "`r Sys.Date()`"
output:
  pdf_document:
  toc: true
toc_depth: 2
df_print: kable
highlight: haddock
latex_engine: xelatex
linkcolor: blue    
urlcolor: blue
citecolor: red
toccolor: blue
---



```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

# load all necessary packages
require(marea)
require(dplyr)
require(ggplot2)
require(stringr)
require(purrr)
require(patchwork)
require(knitr)
require(glue)
require(sf)
require(kableExtra)
require(SSESRGenerator)

```


\newpage


```{r make-map-ess, out.width="75%", fig.align="center", fig.cap="Map of Region: 4X"}

# load spatial data
data("NAFO_shapefiles")
invisible(get(data(list = "coastline", package = "marea")))
bbox <- c(xmin = -75, xmax = -50, ymin = 35, ymax = 50)

target_region <- NAFO_shapefiles %>%
filter(NAFO_1 %in% c("4X"))

# make map
target_area_map <- ggplot() +
  geom_sf(data = target_region, fill = "darkblue") +
  geom_sf(data = coastline) +
  coord_sf(xlim = c(bbox[["xmin"]], bbox[["xmax"]]),
           ylim = c(bbox[["ymin"]], bbox[["ymax"]]),
           expand=F) +
  theme_light()

# print map
target_area_map

```


# Introduction {#introduction}

### About this report

This Ecosystem Summary Report supports DFO's efforts to incorporate ecosystem approaches into Fisheries Science and Resource Management in the Maritimes Region. The purpose of this report is to synthesize ecosystem information to inform an ecosystem approach, as highlighted in the modernized Fisheries Act ([2019; Bill C-68](https://lop.parl.ca/sites/PublicWebsite/default/en_CA/ResearchPublications/LegislativeSummaries/421C68E)). It aligns objectives broadly to those of the [Maritimes Ecosystem Based Management Framework](https://sites.google.com/view/ebminitiative/ebm-initiative-home), and regional/area-based objectives that contribute to Scientific Advice and Management Plans or other Harvest Strategies. The aim of this Ecosystem Summary Report is to provide an integrative suite of ecosystem indicators relevant to a specific fisheries management area that can support Ecosystem Approaches to Fisheries Management ([EAFM](https://www.dfo-mpo.gc.ca/fisheries-peches/initiatives/ecosystems-approach-approche-ecosystemique/index-eng.html)) and Ecosystem Based Fisheries Management ([EBFM](https://sites.google.com/view/maritimesebfmwg/home)) at DFO.

### Report structure

This report outlines ecosystem objectives related to a specific management area (Table 1), and reviews relevant ecosystem indicator trends and the status of the most recent year(s) relative to a reference point (if available) or a long-term average. The results are synthesized to outline potential implications for the management area examined, but not specifically related to a focal stock or species. Where relevant this document provides links to reports and resources for details and methodology. This report summarizes general ecosystem trends to support integration of ecosystem information and expertise into Science Advice (e.g. stock assessment) or Resource Management (e.g. Integrated Fisheries Management Plans), but is intended for a variety of users.

This is a summary of the best available information describing the ecosystem specific to management areas within the Maritimes Region.


\newpage


# Ecosystem Objectives {#ecosystemobjectives}

The Ecosystem objectives outline the ecosystem context and considerations for this management area.

```{r , echo=FALSE, tab.cap="Ecosystem objectives related to EBM Framework Pillar, Management Components, and Indicators." }
# load objectives table
data("ESR_objectives")

ESR_objectives %>%
  as.data.frame() %>%
  kable()

```


\newpage 


# A. Pressures and Stressors {#pressuresandstressors}

## Climate and Oceanography

The management area is strongly influenced by natural variability of the climate system (e.g., Atlantic Multidecadal Oscillation, North Atlantic Oscillation) and, therefore, long-term monitoring is needed to determine how anthropogenic climate change is affecting ocean temperatures and other ocean properties. Changes in climate can directly and indirectly impact stock status by affecting the distribution, seasonal timing, physiology, migration and spawning of marine species.

### Atlantic Multidecadal Oscillation (AMO); global scale

The AMO describes long-duration changes in the sea surface temperature of the North Atlantic Ocean, with cool and warm phases that may last for 20-40 years.

```{r plot-amo, fig.align="center", echo = FALSE,fig.pos="H", fig.width = 5, fig.height=3,fig.cap="The Atlantic Multidecadal Oscillation (AMO) index"}

data("amo")
plot(amo, style = "anomaly") + theme_light(base_size = 8)

```

<!-- add interpretation here -->


\newpage


### North Atlantic Oscillation (NAO); global scale

[Hebert et al. 2024](https://publications.gc.ca/collections/collection_2024/mpo-dfo/Fs97-18-380-eng.pdf) (see Section 2.1)

```{r plot-nao, fig.align="center", echo = FALSE,fig.pos="H", fig.width = 5, fig.height=3,fig.cap="The North Atlantic Oscillation (NAO) index, defined as the winter (December, January, February, March) 500 mb pressure Principal Component Analysis which is representative of the difference between the Icelandic low and Azores high. "}

data("nao")
plot(nao, style = "default") + theme_light(base_size = 8)

```

<!-- add interpretation here -->


\newpage


### AZMP Surface Temperature from Satellite Data; local scale 4XSS

[Casault et al. 2020](https://casaultb.github.io/azmpdata/)

```{r plot-azmp-satellite-temperature, fig.align="center", echo = FALSE,fig.pos="H", fig.width = 5, fig.height=3,fig.cap="Sea surface temperature from satellite data for region 4X"}

# NOTE: azmp_satellite_temperature regions include:
# 4Vn, 4Vs, 4W, 4XeGoM+BoF, 4XeGoMBoF, 4XSS 

# load temperature data
data("azmp_satellite_temperature")

# subset to region
azmp_satellite_temperature_region <-
ea.subset(azmp_satellite_temperature, 'region', c("4XSS"))
rm(azmp_satellite_temperature)
azmp_satellite_temperature_region@meta$region <- "4XSS"

# plot
plot(azmp_satellite_temperature_region, style = "default") + theme_light(base_size = 8)

```

<!-- add interpretation here -->


\newpage


### AZMP Surface Temperature; local scale (Scotian Shelf Box)

[Casault et al. 2020](https://casaultb.github.io/azmpdata/)

```{r plot-azmp-surface-temperature, fig.align="center", echo = FALSE,fig.pos="H", fig.width = 5, fig.height=3,fig.cap="Surface temperature in the Scotian Shelf Region."}

data("azmp_surface_temperature")
plot(azmp_surface_temperature, style = "default") + theme_light(base_size = 8)

```

<!-- add interpretation here -->


\newpage


### AZMP Sea Bottom Temperature; local scale 4X

[Casault et al. 2020](https://casaultb.github.io/azmpdata/)

```{r plot-azmp-bottom-temperature, fig.align="center", echo = FALSE,fig.pos="H", fig.width = 5, fig.height=3,fig.cap="Sea bottom temperature for region 4X"}

# NOTE: azmp_bottom_temperature regions include:
# Cabot Strait, E Georges Bank, Emerald Basin, Georges Basin, Lurcher Shoal, Misaine Bank, 4Vn, 4Vs, 4W, 4XeGoMBoF, 4XSS, 4X 

# load data
data("azmp_bottom_temperature")

# subset to region
azmp_bottom_temperature_region <-
ea.subset(azmp_bottom_temperature, 'region', c("4X"))
rm(azmp_bottom_temperature)
azmp_bottom_temperature_region@meta$region <- "4X"

# plot
plot(azmp_bottom_temperature_region, style = "default") + theme_light(base_size = 8)

```

<!-- add interpretation here -->


\newpage


### AZMP Stratification; local scale (Scotian Shelf Box)

[Layton et al. 2025](https://publications.gc.ca/collections/collection_2025/mpo-dfo/Fs97-18-403-eng.pdf)

```{r plot-azmp-stratification, fig.align="center", echo = FALSE,fig.pos="H", fig.width = 5, fig.height=3,fig.cap="Stratification in the Scotian Shelf Region."}

data("azmp_stratification")
plot(azmp_stratification, style = "default") + theme_light(base_size = 8)

```

<!-- add interpretation here -->


\newpage


### AZMP Salinity; local scale (Scotian Shelf Box)

[Layton et al. 2025](https://publications.gc.ca/collections/collection_2025/mpo-dfo/Fs97-18-403-eng.pdf)

```{r plot-azmp-salinity, fig.align="center", echo = FALSE,fig.pos="H", fig.width = 5, fig.height=3,fig.cap="Salinity in the Scotian Shelf Region."}

data("azmp_salinity")
plot(azmp_salinity, style = "default") + theme_light(base_size = 8)

```

<!-- add interpretation here -->


\newpage


### Ocean Acifification

Template data not available.

```{r plot-ocean-acidification, fig.align="center", echo = FALSE,fig.pos="H", fig.width = 5, fig.height=3,fig.cap=""}

# Add ocean acidifcation data, if available.

```

<!-- add interpretation here -->


\newpage


## Longer-Term Climate Change

### Future Projections and Model Outputs

Template data not available.

```{r plot-future-projections, fig.align="center", echo = FALSE,fig.pos="H", fig.width = 5, fig.height=3,fig.cap=""}

# Add future projections data, if available.

```


\newpage


# B. Ecological {#ecological}

## Commercial Fisheries

### Fisheries Landings; local scale c("4X")

Commercial fishing is an activity that impacts the ecology of a management area through removals of fish from the ecosystem

```{r plot-commercial-fishing, fig.align="center", fig.pos = "H", fig.width = 7, fig.height=4.5,fig.cap="Commercial fisheries landings for target groups in region 4X"}


data("eco_indicators")
# subset to target region
eco_indicators_region <- ea.subset(eco_indicators,"region",c("4X"))
rm(eco_indicators)

plot_commercial_fishing <- function(fishery){
  
  ea_data_fishery <- ea_data(
    data = eco_indicators_region[["data"]][c("year",paste0("landings_",fishery,".L_value"))],
    value_col =  paste0("landings_",fishery,".L_value"),
    data_type =  paste0(str_to_title(fishery), " Landings"),
    region = "4X",
    location_descriptor = "4X NAFO management region",
    units = "tonnes",
    source_citation = eco_indicators_region@meta$source_citation
  )
  
  ea_fishery_plot <- plot(ea_data_fishery, style = "histogram")
  return(ea_fishery_plot)
  
}

# make fisheries plot for each group
fishing_plot_clupeids <- plot_commercial_fishing("CLUPEIDS")
fishing_plot_finfish <- plot_commercial_fishing("FINFISH")
fishing_plot_flatfish  <- plot_commercial_fishing("FLATFISH")
fishing_plot_forage <- plot_commercial_fishing("FORAGE")
fishing_plot_gadoids  <- plot_commercial_fishing("GADOIDS")
fishing_plot_groundfish <- plot_commercial_fishing("GROUNDFISH")
# Note that data for other groups may be available.
# Check eco_indicators columns in marea.

# merge with patchwork
fishing_plot_full <- 
  (fishing_plot_clupeids + 
     fishing_plot_finfish + 
     fishing_plot_flatfish + 
     fishing_plot_forage + 
     fishing_plot_gadoids + 
     fishing_plot_groundfish) &
  theme_light(base_size = 9) 

fishing_plot_full


```

<!-- add interpretation here -->


\newpage


### Fisheries Pressure; local scale c("4X")

This indicator measures the level of exploitation or total fishing pressure at the ecosystem level (Landings/Biomass). Change in this indicator can result from change in biomass, landings or both.

```{r plot-fishing-pressure, fig.align="center", fig.pos = "H", fig.width = 7, fig.height=4.5,fig.cap="Fishing pressure for target groups in region 4X"}


data("eco_indicators")
# subset to target region
eco_indicators_region <- ea.subset(eco_indicators,"region",c("4X"))
rm(eco_indicators)

plot_fishing_pressure <- function(fishery){
  
  ea_data_fp <- ea_data(
    data = eco_indicators_region[["data"]][c("year",paste0("FP_",fishery,"_value"))],
    value_col =  paste0("FP_",fishery,"_value"),
    data_type = paste0(str_to_title(fishery), " FP"),
    region = "4X",
    location_descriptor = "4X NAFO management region",
    units = "",
    source_citation = eco_indicators_region@meta$source_citation
  )
  
  ea_fp_plot <- plot(ea_data_fp, style = "plain")
  return(ea_fp_plot)
  
}

# make fisheries plot for each group
fp_plot_clupeids <- plot_fishing_pressure("CLUPEIDS")
fp_plot_finfish <- plot_fishing_pressure("FINFISH")
fp_plot_flatfish  <- plot_fishing_pressure("FLATFISH")
fp_plot_forage <- plot_fishing_pressure("FORAGE")
fp_plot_gadoids  <- plot_fishing_pressure("GADOIDS")
fp_plot_groundfish <- plot_fishing_pressure("GROUNDFISH")
# Note that data for other groups may be available.
# Check eco_indicators columns in marea.

# merge with patchwork
fp_plot_full <- 
  (fp_plot_clupeids + 
     fp_plot_finfish + 
     fp_plot_flatfish + 
     fp_plot_forage + 
     fp_plot_gadoids + 
     fp_plot_groundfish) &
  theme_light(base_size = 9) 

fp_plot_full


```

<!-- add interpretation here -->


\newpage


## Habitat

### Cumulative Impacts

Template data not available.

```{r plot-cumulative-impacts, fig.align="center", echo = FALSE,fig.pos="H", fig.width = 5, fig.height=3,fig.cap=""}

# Add Cumulative impacts data, if available.

```

<!-- add interpretation here -->


\newpage


## Productivity

### Nitrate

Template data not available.

```{r plot-nitrate, fig.align="center", echo = FALSE,fig.pos="H", fig.width = 5, fig.height=3,fig.cap=""}

# Add Nitrate data, if available.

```

<!-- add interpretation here -->


\newpage


### Bloom Statistics

Template data not available.

```{r plot-bloom-stat, fig.align="center", echo = FALSE,fig.pos="H", fig.width = 5, fig.height=3,fig.cap=""}

# Add Bloom Statistic data, if available.

```

<!-- add interpretation here -->


\newpage


### Zooplankton Biomass

Template data not available.

```{r plot-zoo-biomass, fig.align="center", echo = FALSE,fig.pos="H", fig.width = 5, fig.height=3,fig.cap=""}

# Add Zooplankton Biomass data, if available.

```

<!-- add interpretation here -->


\newpage


### Large Fish Indicator

Template data not available.

```{r plot-large-fish, fig.align="center", echo = FALSE,fig.pos="H", fig.width = 5, fig.height=3,fig.cap=""}

# Add Large Fish Indicator data, if available.

```

<!-- add interpretation here -->


\newpage


## Biodiversity

### Shannon Diversity Index, Margalef Richness, Heip’s Evenness; local scale c("4X")

Shannon Diversity Index accounts for both abundance and evenness of species present in a community. Margalef Richness measures the number of species present accounting for sampling effects. Heips Evenness measures how equally the species richness contributes to the total abundance or biomass of the community. Due to improvements in species identification over time, long-term Shannon Diversity and Margalef richness can be misleading, only years 2000-2020 were used here.

```{r plot-biodiversity-vars, fig.align="center", fig.pos = "H", fig.width = 7, fig.height=2.5,fig.cap="Biodiversity Variables for target groups in region 4X"}

data("eco_indicators")
# subset to target region
eco_indicators_region <- ea.subset(eco_indicators,"region",c("4X"))
rm(eco_indicators)

plot_biodiversity_vars <- function(var_col, var_name){
  
  ea_data_biodiversity <- ea_data(
    data = eco_indicators_region[["data"]][c("year",var_col)],
    value_col =  var_col,
    data_type = var_name,
    region = "ESS",
    location_descriptor = "ESS NAFO management region",
    units = "",
    source_citation = eco_indicators_region@meta$source_citation
  )
  
  ea_biodiversity_plot <- plot(ea_data_biodiversity, style = "plain")
  return(ea_biodiversity_plot)
  
}

bio_plot_shannon <- plot_biodiversity_vars("ShannonDiversity_ALL_value","Shannon Diversity")
bio_plot_margalef <- plot_biodiversity_vars("MargalefRichness_ALL_value","Margalef Richness")
bio_plot_heips <- plot_biodiversity_vars("Heips_ALL_value","Heips Evenness")


# merge with patchwork
bio_plot_full <- 
  (bio_plot_shannon + 
     bio_plot_margalef + 
     bio_plot_heips) &
  theme_light(base_size = 9) 

bio_plot_full


```

<!-- add interpretation here -->


\newpage


### Guild-Level Biomass; local scale c("4X")

Guild-level biomasses address structural attributes of food webs, and can also serve as proxies for ecosystem functioning ([Tam et al. 2017](https://academic.oup.com/icesjms/article/74/7/2040/2970046)). This indicator includes multiple guilds for fish and invertebrates from the RV Survey.

```{r plot-guild-biomass, fig.align="center", fig.pos = "H", fig.width = 7, fig.height=4.5,fig.cap="Biomass of target groups from RV surveys in region 4X"}

data("eco_indicators")
# subset to target region
eco_indicators_region <- ea.subset(eco_indicators,"region",c("4X"))
rm(eco_indicators)

plot_guild_biomass <- function(guild){
  
  ea_data_guild <- ea_data(
    data = eco_indicators_region[["data"]][c("year",paste0("BIOMASS_",guild,"_value"))],
    value_col =  paste0("BIOMASS_",guild,"_value"),
    data_type = paste0(str_to_title(guild), " Biomass"),
    region = "4X",
    location_descriptor = "4X NAFO management region",
    units = "tonnes",
    source_citation = eco_indicators_region@meta$source_citation
  )
  
  ea_guild_biomass_plot <- plot(ea_data_guild, style = "default")
  return(ea_guild_biomass_plot)
  
}

# make fisheries plot for each group
biomass_plot_clupeids <- plot_guild_biomass("CLUPEIDS")
biomass_plot_finfish <- plot_guild_biomass("FINFISH")
biomass_plot_flatfish  <- plot_guild_biomass("FLATFISH")
biomass_plot_forage <- plot_guild_biomass("FORAGE")
biomass_plot_gadoids  <- plot_guild_biomass("GADOIDS")
biomass_plot_groundfish <- plot_guild_biomass("GROUNDFISH")
# Note that data for other groups may be available. 
# Check eco_indicators columns in marea.

# merge with patchwork
biomass_plot_full <- 
  (biomass_plot_clupeids + 
     biomass_plot_finfish + 
     biomass_plot_flatfish + 
     biomass_plot_forage + 
     biomass_plot_gadoids + 
     biomass_plot_groundfish) &
  theme_light(base_size = 9) 

biomass_plot_full


```

<!-- add interpretation here -->


\newpage


## Non-Target Species

### Non-Target Species

Template data not available.

```{r plot-non-target-species, fig.align="center", echo = FALSE,fig.pos="H", fig.width = 5, fig.height=3,fig.cap=""}

# Add Non-target species data, if available.

```

<!-- add interpretation here -->


\newpage


# C. Economic {#economic}

## Commercial Fisheries

Commercial fishing is an important activity in the Maritimes Region that provides both cultural and economic value to coastal communities.

### Landing Value by Species Category

Template data not available.

```{r plot-commercial-fisheries-value, fig.align="center", echo = FALSE,fig.pos="H", fig.width = 5, fig.height=3,fig.cap=""}

# Add Commercial Fisheries data, if available.

```

<!-- add interpretation here -->


\newpage


# D. Social-Cultural {#social-cultural}

Currently unavailable


\newpage


# E. Governance {#governance}

Currently unavailable


\newpage


